Ecological management of stochastic systems with long transients

### 2020-06-22 manuscript/manuscript.Rmd
EM adds line numbers already
revisions WIP

---
title: "Ecological management of stochastic systems with long transients"
date: "`r Sys.Date()`"
author:
  - name: Carl Boettiger
    email: cboettig@berkeley.edu
    affiliation: ucb
#    footnote: Corresponding Author
address:
  - code: ucb
    address: "ESPM Department, University of California, 130 Mulford Hall Berkeley, CA 94720-3114, USA"
abstract: |
  Long transients may be common in ecological dynamics, but the implications such dynamics have for ecological management have not been fully explored.  Long transient periods can easily be mistaken for stable-state dynamics, but may require dramatically different management policies.  Here, I explore the optimal management of stochastic ecological systems that may contain either a tipping point or a ghost attractor: a recently recognized mechanism which can give rise to long transients. I consider three approaches of increasing sophistication: (1) dynamic management under a fixed model, (2) management that accounts for uncertainty over possible models, and (3) adaptive management that actively learns the correct model over the management process. This analysis confirms the prediction that long transients can create considerable uncertainty and give rise to very different optimal management policies, but also illustrates that dynamic management that can either plan for this uncertainty or actively learn to decrease the uncertainty can promote successful management of long transients.
journal: Theoretical ecology
layout: 3p
bibliography: refs.bib
output: rticles::elsevier_article
keywords: 
  - transients, 
  - optimal control, 
  - adaptive management, 
  - stochasticity, 
  - uncertainty, 
  - ecological management
preamble: |
  \journal{Theoretical Ecology}
  \usepackage{lineno}
  \linenumbers

@@

I consider a generic model of population dynamics which can exhibit phenomena such as bistability and a ghost attractor dynamics.  This model is amenable to interpretation in many ecological contexts (sensu @Levins1966 sacrificing precision to realism and generality), such as the collapse of a fish population, recovery of an endangered species, or the outbreak a pest.  To provide a more tangible context to frame our questions of ecological management, I will focus on the latter scenario of a pest outbreak.  In this scenario, we imagine a species such as the Pine bark beetle, (e.g. *Dendroctonus ponderosae*, @Harvey2014) whose recent boom in population has devastated much of Northern America's pine forests.  The model is a stochastic version of the consumer-resource model of @May1977,

@@

Where $X_t$ is the population density of the focal species (the beetle, in our case).  The growth  rate of the species is determined by an intrinsic growth rate $r$ and carrying capacity $K$. The species is under mortality pressure from another species (the natural defenses of trees, in our context) which has an S curve saturation which reaches its half maximum at $H$ and maximum rate of $a$. Parameter $Q$ determines the steepness of the saturating response. The dynamics are subject to environmental white noise proportional to the population size, $\xi_t X_t$, uncorrelated Gaussian noise $\xi \sim \mathcal{N}(0,\sigma)$ with mean zero and standard deviation $\sigma$ [@Boettiger2018]. Model dynamics are illustrated in Fig 1 under two different parameter choices for the saturation response $a$: one which corresponds to a ghost attractor and one to a bistable system.

@@

```{r figure1, fig.cap="Change in population size $X_{t+1} - X_t = f(X_t)$. Equilibria occur when the change curve crosses zero, stable points arise at crossings with negative slopes, unstable equilibria have positive slopes.  By changing parameter $a$, the system goes from bistable (red, $a = 28$) to a ghost attractor (blue, $a = 27.2$). For both curves, $r = 0.8$, $K = 153$, $Q = 2$, $H = 20$."}

@@

Models with this structure are also capable of producing long transient dynamics instead of alternative stable state dynamics by slightly changing the chosen parameter values.  Observe that by decreasing $a$, the lower stable point vanishes, but the rate of change of the system (e.g. the vertical axis, Fig 1) is still small in magnitude, and thus dynamics near this region move slowly.  This is a simple example of what @Hastings2018 has dubbed a "Ghost Attractor" -- the attractor vanishes at the fold bifurcation, but it's effects can still be felt and observed.  This could result from a weakened ability of the trees to respond to attack, which could result from additional stress such as prolonged drought or fire, both potentially intensified under changing climate.

>>>>

---
title: "Ecological management of stochastic systems with long transients"
date: "`r Sys.Date()`"
author:
  - name: Carl Boettiger
    email: cboettig@berkeley.edu
    affiliation: ucb
#    footnote: Corresponding Author
address:
  - code: ucb
    address: "ESPM Department, University of California, 130 Mulford Hall Berkeley, CA 94720-3114, USA"
abstract: |
  Long transients may be common in ecological dynamics, but the implications such dynamics have for ecological management have not been fully explored.  Long transient periods can easily be mistaken for stable-state dynamics, but may require dramatically different management policies.  Here, I explore the optimal management of stochastic ecological systems that may contain either a tipping point or a ghost attractor: an important mechanism which can give rise to long transients. I consider three approaches of increasing sophistication: (1) dynamic management under a fixed model, (2) management that accounts for uncertainty over possible models, and (3) adaptive management that actively learns the correct model over the management process. This analysis confirms the prediction that long transients can create considerable uncertainty and give rise to very different optimal management policies, but also illustrates that dynamic management that can either plan for this uncertainty or actively learn to decrease the uncertainty can promote successful management of long transients.
journal: Theoretical ecology
layout: 3p
bibliography: refs.bib
output: rticles::elsevier_article
keywords: 
  - transients, 
  - optimal control, 
  - adaptive management, 
  - stochasticity, 
  - uncertainty, 
  - ecological management
preamble: |
  \journal{Theoretical Ecology}

## EditorialManager adds its own linenumbers to pdf already
#\usepackage{lineno}
# \linenumbers

@@

I consider a generic model of population dynamics which can exhibit phenomena such as bistability and a ghost attractor dynamics.  This model is amenable to interpretation in many ecological contexts (sensu @Levins1966 sacrificing precision to realism and generality), such as the collapse of a fish population, recovery of an endangered species, or the outbreak a pest.  To provide a more tangible context to frame our questions of ecological management, I will focus on the latter scenario of a pest outbreak.  In this scenario, we imagine a species such as the Pine bark beetle, (e.g. *Dendroctonus ponderosae*, @Harvey2014) whose recent boom in population has devastated much of Northern America's pine forests.  The model is a stochastic extension of the deterministic consumer-resource model of @May1977,

@@

Where $X_t$ is the population density of the focal species (the beetle, in our case).  The growth  rate of the species is determined by an intrinsic growth rate $r$ and carrying capacity $K$. The species is under mortality pressure from another species (the natural defenses of trees, in our context) which has an S curve saturation which reaches its half maximum at $H$ and maximum rate of $a$. Parameter $Q$ determines the steepness of the saturating response. The stochastic formulation follows directly from interpreting May's model as the average dynamics of a process subject to to environmental white noise proportional to the population size, $\xi_t X_t$, uncorrelated Gaussian noise $\xi \sim \mathcal{N}(0,\sigma)$ with mean zero and standard deviation $\sigma$ [@Boettiger2018]. Model dynamics are illustrated in Fig 1 under two different parameter choices for the saturation response $a$: one which corresponds to a ghost attractor and one to a bistable system.

@@

```{r figure1, fig.cap="Change in population size $X_{t+1} - X_t = f(X_t)$. Equilibria occur when the change curve crosses zero, stable points arise at crossings with negative slopes, unstable equilibria have positive slopes.  By changing parameter $a$, the system goes from bistable (blue, $a = 28$) to a ghost attractor (red, $a = 27.2$). For both curves, $r = 0.8$, $K = 153$, $Q = 2$, $H = 20$."}

@@

Models with this structure are also capable of producing long transient dynamics instead of alternative stable state dynamics by slightly changing the chosen parameter values.  Observe that by decreasing $a$, the lower stable point vanishes, but the rate of change of the system (e.g. the vertical axis, Fig 1) is still small in magnitude, and thus dynamics near this region move slowly.  This is a simple example of a "Ghost Attractor" [@Deco2012] -- the attractor vanishes at the fold bifurcation, but it's effects can still be felt and observed [@Grossman1980; @Hastings2018].  This could result from a weakened ability of the trees to respond to attack, which could result from additional stress such as prolonged drought or fire, both potentially intensified under changing climate.
